Hybrid MILP to efficiently and accuratly \\ solve hard DNN verification instances

### 2024-09-22 ICLR25/introduction.tex
valabstraction

Our three main contributions address the challenges of verifying {\em hard} DNNs efficiently:
\begin{enumerate}
	\item  Our first contribution is to study the LP relaxation for the complete MILP
	encoding of ReLU-DNNs. We establish in Proposition \ref{LP} an explicit abstraction set corresponding to the LP constraints.
	\item This allows us to propose a measure of utility for each neuron, providing an upper bound on the improvement on accuracy made if we treat this neuron with an integer variable instead of the LP relaxation. Choosing the neurons with the top utility and encoding them with integer variables gives rise to a partial MILP which is much more efficient than previous attempts \cite{9211410}, necessitating much less integer variables for the same accuracy.

>>>>

Our three main contributions address the challenges of verifying {\em hard} DNNs efficiently:
\begin{enumerate}
	\item  Our first contribution is to study the LP relaxation for the complete MILP
	encoding of ReLU-DNNs. We establish in Proposition \ref{LP} an explicit abstraction set corresponding to the LP constraints.
	\item This allows us to propose a measure of utility for each neuron, providing an upper bound on the improvement on accuracy made if we treat this neuron with an integer variable instead of the LP relaxation. Choosing the neurons with the top utility and encoding them with integer variables gives rise to a partial MILP which is much more efficient than previous attempts \cite{DivideAndSlide}, necessitating much less integer variables for the same accuracy.

### 2024-09-22 ICLR25/relatedwork.tex
valabstraction

We compare {\CMP} with major verification tools for DNNs to clarify our methodology and its distinction from the existing state-of-the-art. CMP scales while preserving good accuracy, through targeting a limited number of binary variables based on compensating strength, stricking a good balance between exact encoding of a DNN using MILP~\cite{MILP} (too slow) and LP relaxation (too inaccurate). MIPplanet~\cite{MIPplanet} opts for a different selection of binary variables, and execute one large MILP encoding instead of CMP's many small encodings, which significantly reduce the number of binary variables necessary for each encoding. In \cite{9211410}, small encodings are also considered, however with a straightforward choice of binary nodes based on the weight of outgoing edges, which need much more integer variables (thus runtime) to reach the same accuracy as our utility function (see Table \ref{tab:example1}), and restrict integer variables to the previous layer, which limits the accuracy reachable.

>>>>

We compare {\CMP} with major verification tools for DNNs to clarify our methodology and its distinction from the existing state-of-the-art. CMP scales while preserving good accuracy, through targeting a limited number of binary variables based on compensating strength, stricking a good balance between exact encoding of a DNN using MILP~\cite{MILP} (too slow) and LP relaxation (too inaccurate). MIPplanet~\cite{MIPplanet} opts for a different selection of binary variables, and execute one large MILP encoding instead of CMP's many small encodings, which significantly reduce the number of binary variables necessary for each encoding. In \cite{DivideAndSlide}, small encodings are also considered, however with a straightforward choice of binary nodes based on the weight of outgoing edges, which need much more integer variables (thus runtime) to reach the same accuracy as our utility function (see Table \ref{tab:example1}), and restrict integer variables to the previous layer, which limits the accuracy reachable.

### 2024-09-22 ICLR25/valabstraction.tex
Add paper in new format
valabstraction

\subsection{The Box and DeepPoly Abstractions}

\definecolor{applegreen}{rgb}{0.55, 0.71, 0.0}

\begin{figure}[b!]
	\centering
	\begin{tikzpicture}

@@

		\node[circle, draw= purple, thick, minimum width = 20,
		minimum height = 20] (input2) at ($(input1) + (0,-1.5)$) {$n_2$};

@@

		\node[circle, draw= blue, thick, minimum width = 20,
		minimum height = 20] (hidden5) at ($(input1) + (7.5,0)$){$n_5$};
		\node[circle, draw= black, thick] (hidden6) at ($(input1) + (7.5,-1.5)$) {$n_6$};

@@

		\draw[->,color=applegreen, thick] (input2) -- (hidden1) node[near start, below] {$1$};
		\draw[->,color=red, thick] (input2) -- (hidden2)node[near start, below] {$-1$};

		\draw[->,color=applegreen, thick] (hidden1) -- (hidden3) node[midway, above] {$\max(n_1,0)$};
		\draw[->,color=red, thick] (hidden2) -- (hidden4) node[midway, above] {$\max(n_2,0)$};

		\draw[->,color=applegreen, thick] (hidden3) -- (hidden5) node[near start, above] {$1$};			
		\draw[->,thick] (hidden3) -- (hidden6) node[near start, above] {$0$};

		\draw[->,color=red, thick] (hidden4) -- (hidden5)node[near start, below] {$2$};
		\draw[->,thick] (hidden4) -- (hidden6)node[near start, below] {$1$};

@@

	\end{tikzpicture}
	\caption{A DNN example from \cite{kpoly}: every neuron is separated into 2 nodes, 
	$n$ pre- and $\hat{n}$ post-ReLU activation function. The pair of paths 
	$({\color{applegreen}n_2 n_3 n_5},{\color{red}n_2 n_4 n_5})$ is {\em compensating} 
	(weights ${\color{applegreen}1},{\color{red}-2}$).}
	\label{fig1}
\end{figure}

Firstly, it's important to note that weighted sums represent a linear function, which can be explicitly expressed with relative ease. However, the ReLU (Rectified Linear Unit) function presents a challenge in terms of accurate representation. Although ReLU is a relatively straightforward piecewise linear function with two modes (one for $x<0$ and another for $x \geq 0$), it is not linear. The complexity arises when considering the compounded effects of the ReLU function across the various layers of a ReLU DNN. It's worth noting that representing $\ReLU(x)$ precisely is feasible when $x$ is "stable," meaning it's consistently positive or consistently negative, as there's only one linear mode involved in each scenario. Consequently, the primary challenge lies in addressing "unstable" neurons, where the linearity of the function does not hold consistently.

Consider the simpler abstraction, termed ``Box abstraction" \cite{deeppoly}: it inductively computes the bounds for each neuron in the subsequent layer independently. This is achieved by considering the weighted sum of the bounds from the previous layer, followed by clipping the lower bound at $\max(0,$ lower bound$)$ to represent the ReLU function, and so forth. 
For all $i \geq 3$, define $x_i=\val_{\vx}(n_i)$, where $\vx=(x_1,x_2)$.

Taking the DNN example from Fig \ref{fig1}, assume $x_1,x_2 \in [-1,1]$. This implies that $x_3,x_4 \in [-2,2]$. After applying the ReLU function, $\hat{x}_3,\hat{x}_4$ are constrained to $[0,2]$, leading to $x_5 \in [0,6]$ and $x_6 \in [0,2]$. 
The bounds for $n_1, \ldots, n_4$ are exact, meaning for every $\alpha$ within the range, an input $\vy$ can be found such that $\val_{\vy}(n_i)=\alpha$. However, this precision is lost from the next layer (beginning with $n_5, n_6$) due to potential dependencies among preceding neurons. For example, it is impossible for $x_5=\Val_{\vx}(n_5)$ to reach $6$, as it would necessitate both $x_3=2$ and $x_4=2$, which is not possible at the same time as 
$x_3=2$ implies $x_1=x_2=1$ and $x_4=2$ implies $x_2=-1$ (and $x_1=1$), a contradiction.

An extremely efficient algorithm that alleviates some of the inaccuracies is DeepPoly \cite{deeppoly}, also independently discovered as the CROWN algorithm \cite{crown}. Instead of fixed bounds for each neuron $n$ in layer $k$, DeepPoly maintains two affine functions representing the lower and upper bounds of the neuron's value, based on inputs from the previous layer $k-1$. For instance, denote $f_i \leq x_i \leq g_i$ with, for example, $f_{3}(x_3)=f_4(x_4)=0$ and $g_3(x_3) = \frac{x_3+2}{2}$, $g_4(x_4) = \frac{x_4+2}{2}$. This leads to $x_5 \leq g_3(x_3) + 2 g_4(x_4) = \frac{x_3 + 2x_4 + 6}{2} = \frac{3x_1 - x_2 + 6}{2}\leq 5$.

For bounds $[\alpha,\beta]$ on $x_i$, the optimal linear function for the upper bound is $g_i(x_i)= \beta \frac{x_i-\alpha}{\beta-\alpha}$ for ReLU nodes. There are two options for the lower bound: $f^1_i(x_i) = 0$ or $f^2_i(x_i)=x_i$. DeepPoly selects between these based on the values of $\alpha$ and $\beta$ (for unstable neurons): if $|\alpha|\geq |\beta|$, then $f_i=f^1_i$ is chosen, otherwise, for $|\beta|>|\alpha|$, $f_i=f^2_i$ is selected. The variation, {\em $\overline{\mbox{DeepPoly}}$}, consistently chooses $f^1_i$, and never $f^2_i$. Contrary to DeepPoly, {\em $\overline{\mbox{DeepPoly}}$} encompasses the ``Box abstraction." For instance, with bounds $[-0.2,5]$ on $x_i$, 
standard DeepPoly would conclude $\ReLU(x_i) \in [-0.2,5]$, while 
{\em $\overline{\mbox{DeepPoly}}$} would deduce $\ReLU(x_i) \in [0,5]$ similarly as the Box abstraction.

\iffalse
\subsection{PRIMA and $\beta$-CROWN}
\fi

\subsection{MILP and LP encodings for DNNs}

@@

>>>>

\subsection{The Box Abstractions}

%\definecolor{applegreen}{rgb}{0.55, 0.71, 0.0}

Firstly, it's important to note that weighted sums represent a linear function, which can be explicitly expressed with relative ease. However, the ReLU (Rectified Linear Unit) function presents a challenge in terms of accurate representation. Although ReLU is a relatively straightforward piecewise linear function with two modes (one for $x<0$ and another for $x \geq 0$), it is not linear. The complexity arises when considering the compounded effects of the ReLU function across the various layers of a ReLU DNN. It's worth noting that representing $\ReLU(x)$ precisely is feasible when $x$ is "{\em stable}", meaning it's consistently positive or consistently negative, as there's only one linear mode involved in each scenario. Consequently, the primary challenge lies in addressing "{\em unstable}" neurons, where the linearity of the function does not hold consistently.

\begin{figure}[t!]
	\centering
	\begin{tikzpicture}

@@

		\node[circle, draw= black, thick, minimum width = 20,
		minimum height = 20] (input2) at ($(input1) + (0,-1.5)$) {$n_2$};

@@

		\node[circle, draw= black, thick, minimum width = 20,
		minimum height = 20] (hidden5) at ($(input1) + (7.5,0)$){$n_5$};
		\node[circle, draw= black, thick] (hidden6) at ($(input1) + (7.5,-1.5)$) {$n_6$};

@@

		\draw[->, thick] (input2) -- (hidden1) node[near start, below] {$1$};
		\draw[->, thick] (input2) -- (hidden2)node[near start, below] {$-1$};

		\draw[->, thick] (hidden1) -- (hidden3) node[midway, above] {$\max(n_1,0)$};
		\draw[->, thick] (hidden2) -- (hidden4) node[midway, above] {$\max(n_2,0)$};

		\draw[->,thick] (hidden3) -- (hidden5) node[near start, above] {$1$};			
		\draw[->,thick] (hidden3) -- (hidden6) node[near start, above] {$0$};

		\draw[->,thick] (hidden4) -- (hidden5)node[near start, below] {$2$};
		\draw[->,thick] (hidden4) -- (hidden6)node[near start, below] {$1$};

@@

	\end{tikzpicture}
	\caption{A DNN. Every neuron is separated into 2 nodes, 
	$n$ pre- and $\hat{n}$ post-ReLU activation.} 
	%The pair of paths 	$({\color{applegreen}n_2 n_3 n_5},{\color{red}n_2 n_4 n_5})$ is {\em compensating} 	(weights ${\color{applegreen}1},{\color{red}-2}$).}
	\label{fig1}
\end{figure}

Consider the simpler abstraction, termed ``Box abstraction" \cite{deeppoly}: it inductively computes the bounds for each neuron in the subsequent layer independently. This is achieved by considering the weighted sum of the bounds from the previous layer, followed by clipping the lower bound at $\max(0,$ lower bound$)$ to represent the ReLU function, and so forth. 
For all $i \geq 3$, define $x_i=\val_{\vx}(n_i)$, where $\vx=(x_1,x_2)$.
Taking the DNN example from Fig \ref{fig1}, assume $x_1,x_2 \in [-1,1]$. This implies that $x_3,x_4 \in [-2,2]$. After applying the ReLU function, $\hat{x}_3,\hat{x}_4$ are constrained to $[0,2]$, leading to $x_5 \in [0,6]$ and $x_6 \in [0,2]$. 
The bounds for $n_1, \ldots, n_4$ are exact, meaning for every $\alpha$ within the range, an input $\vy$ can be found such that $\val_{\vy}(n_i)=\alpha$. However, this precision is lost from the next layer (beginning with $n_5, n_6$) due to potential dependencies among preceding neurons. For example, it is impossible for $x_5=\Val_{\vx}(n_5)$ to reach $6$, as it would necessitate both $x_3=2$ and $x_4=2$, which is not possible at the same time as 
$x_3=2$ implies $x_1=x_2=1$ and $x_4=2$ implies $x_2=-1$ (and $x_1=1$), a contradiction.

\iffalse
An extremely efficient algorithm that alleviates some of the inaccuracies is DeepPoly \cite{deeppoly}, also independently discovered as the CROWN algorithm \cite{crown}. Instead of fixed bounds for each neuron $n$ in layer $k$, DeepPoly maintains two affine functions representing the lower and upper bounds of the neuron's value, based on inputs from the previous layer $k-1$. For instance, denote $f_i \leq x_i \leq g_i$ with, for example, $f_{3}(x_3)=f_4(x_4)=0$ and $g_3(x_3) = \frac{x_3+2}{2}$, $g_4(x_4) = \frac{x_4+2}{2}$. This leads to $x_5 \leq g_3(x_3) + 2 g_4(x_4) = \frac{x_3 + 2x_4 + 6}{2} = \frac{3x_1 - x_2 + 6}{2}\leq 5$. For bounds $[\alpha,\beta]$ on $x_i$, the optimal linear function for the upper bound is $g_i(x_i)= \beta \frac{x_i-\alpha}{\beta-\alpha}$ for ReLU nodes. There are two options for the lower bound: $f^1_i(x_i) = 0$ or $f^2_i(x_i)=x_i$. DeepPoly selects between these based on the values of $\alpha$ and $\beta$ (for unstable neurons): if $|\alpha|\geq |\beta|$, then $f_i=f^1_i$ is chosen, otherwise, for $|\beta|>|\alpha|$, $f_i=f^2_i$ is selected. 
\fi

\subsection{MILP, LP and partial MILP encodings for DNNs}

@@

In this paper, we propose to use {\em partial MILP}, to get interesting trade-offs between accuracy and runtime: 
for a set of unstable neurons $X$, we denote by MILP$_X$ the MILP encoding where variables encoding $X$ are binary, and other variables are linear variables using the LP relaxation. We say that nodes in $X$ are {\em opened}. 
To further limit the number of binary variables needed for a given accuracy, we devise the same iterative approach as the box abstraction or DeepPoly \cite{deeppoly}, computing lower 
and upper bounds $\LB(n),\UB(n)$ for neurons $n$ of a layer, that are used when computing values of the next layer, thus necessitating less variables from previous layers.

The crucial factor in such an approach is to {\em select} few opened neurons in $X$ which are the most important for the accuracy. An extreme strategy was adopted in \cite{DivideAndSlide}, where only nodes of the immediate previous layer could be opened, and the measure to choose a neuron $a$ to open to compute the values for neuron $b$ was to consider $|W_{ab}| (UB(a)-LB(a))$. To obtain a more accurate measure, that can also scale to deeper layers and is not limited to open nodes from the immediate previous layer, we first study the LP abstraction.
